On $K_2$-analogue of Serre problem for $\SL_4$

### 2018-05-01 paper.tex
Added construction of sigma maps

\documentclass[oneside, 10pt]{amsart}
\usepackage[breaklinks=true]{hyperref}
\usepackage[hyperref=true, backend=bibtex, firstinits=true, citestyle=numeric-comp, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}
\usepackage[capitalize]{cleveref}
\usepackage[matrix,arrow,curve]{xy}
\usepackage{enumitem}

\addbibresource{lemma.bib}
\renewbibmacro*{volume+number+eid}{\ifentrytype{article}{\- \iffieldundef{volume}{}{Vol.~\printfield{volume},}\iffieldundef{number}{}{ No.~\printfield{number},}}}
\renewbibmacro{in:}{\ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
\newbibmacro{string+doi}[1]{\iffieldundef{doi}{\iffieldundef{url}{#1}{\href{\thefield{url}}{#1}}}{\href{http://dx.doi.org/\thefield{doi}}{#1}}}
\DeclareFieldFormat[article, inproceedings, inbook, book, thesis]{title}{\usebibmacro{string+doi}{\mkbibquote{#1}}}
\renewcommand*{\bibfont}{\footnotesize}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\rk}{rk}
\newcommand{\XX}[1]{\mathcal{X}_{#1}}
\newcommand{\RR}[1]{\mathcal{R}_{#1}}

@@

\newtheorem{lem}{Lemma}
\numberwithin{lem}{section}
\Crefname{lemma}{Lemma}{Lemmas}

\newtheorem{cor}[lem]{Corollary}
\Crefname{cor}{Corollary}{Corollaries}

\newtheorem{prop}[lem]{Proposition}
\Crefname{prop}{Proposition}{Propositions}

@@

%\newtheorem{def}[lem]{Definition}
\Crefname{def}{Definition}{Definitions}
\newtheorem{example}[lem]{Example}
\Crefname{example}{Example}{Examples}

@@

\newtheorem{rem}[lem]{Remark}
\Crefname{rem}{Remark}{Remarks}

@@

\subsection{Construction of \texorpdfstring{$\sigma$}{sigma}-morphisms for $SL_4$}

>>>>

\documentclass[oneside, 10pt]{amsart}
\usepackage{amscd, amsmath, amssymb, amsthm, amsfonts, amstext, verbatim, mathtools, xfrac, microtype, nameref, thmtools}
\usepackage[breaklinks=true]{hyperref}
\usepackage[hyperref=true, backend=bibtex, firstinits=true, citestyle=numeric-comp, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}
\usepackage[capitalize]{cleveref}
\usepackage[matrix,arrow,curve]{xy}
\usepackage{enumitem}

\addbibresource{paper.bib}
\renewbibmacro*{volume+number+eid}{\ifentrytype{article}{\- \iffieldundef{volume}{}{Vol.~\printfield{volume},}\iffieldundef{number}{}{ No.~\printfield{number},}}}
\renewbibmacro{in:}{\ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
\newbibmacro{string+doi}[1]{\iffieldundef{doi}{\iffieldundef{url}{#1}{\href{\thefield{url}}{#1}}}{\href{http://dx.doi.org/\thefield{doi}}{#1}}}
\DeclareFieldFormat[article, inproceedings, inbook, book, thesis]{title}{\usebibmacro{string+doi}{\mkbibquote{#1}}}
\renewcommand*{\bibfont}{\footnotesize}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\T}{T}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\Aut}{Aut}
\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\rk}{rk}

\newcommand{\XX}[1]{\mathcal{X}_{#1}}
\newcommand{\RR}[1]{\mathcal{R}_{#1}}

@@

\newtheorem{lemma}{Lemma}
\numberwithin{lemma}{section}
\Crefname{lemma}{Lemma}{Lemmas}
\newlist{lemlist}{enumerate}{1} \setlist[lemlist]{label={\rm(\arabic{lemlisti})}, ref=\thelemma.(\arabic{lemlisti}),noitemsep} \Crefname{lemlisti}{Lemma}{Lemma}

\newtheorem{cor}[lemma]{Corollary}
\Crefname{cor}{Corollary}{Corollaries}

\newtheorem{prop}[lemma]{Proposition}
\Crefname{prop}{Proposition}{Propositions}

@@

%\newtheorem{def}[lem]{Definition}
\Crefname{def}{Definition}{Definitions}
\newtheorem{example}[lemma]{Example}
\Crefname{example}{Example}{Examples}

@@

\newtheorem{rem}[lemma]{Remark}
\Crefname{rem}{Remark}{Remarks}

@@

\subsection{Construction of \texorpdfstring{$\sigma$}{sigma}-morphisms for $SL_4$}
Let $R$ be a local ring and assume that $n\geq 4$.
For $1\leq k\leq n$ denote by $\delta_k$ the $n\times n$ matrix of $\GL(n, R[t, t^{-1}])$ which differs from the unit matrix 
 only by the element $t$ on the $k$-th place of its diagonal.
Recall from~\cite[Corollary~4]{Ka77} that there exists unique automorphisms $\beta_{\delta_k}$, $\beta_{\delta_k^{-1}}$ of $\St(n, R[t, t^{-1}])$
 modeling conjugation with matrices $\delta_k$ and $\delta_k^{-1}$ (see diagram~\eqref{eq:sigma-diagram} below)

Our immediate goal is to define analogues of these automorphisms on the level of Steinberg group $\St(n, R[t])$.
First of all, let us introduce the subgroups of $\St(n, R[t])$ on which these morphisms will be defined.
Denote by $\mathcal{X}_k$ (resp. $\mathcal{X}^k$) the set consisting of those generators $x_{ij}(\xi)$ of $\St(n, R[t])$ for which 
 either $i\neq k$ (resp. $j\neq k$) or $\xi\in R[t]$ is divisible by $t$.
Denote by $N_k$ and $N^k$ the subgroups of $\St(n, R[t])$ generated by $\mathcal{X}_k$ and $\mathcal{X}^k$, respectively.
Our goal is to construct mutually inverse morphisms $\sigma_k$, $\sigma_k^{-1}$ fitting into the following diagram.
\begin{equation} \label{eq:sigma-diagram}
\xymatrix{ N_k\ar@{-->}[r]^{\sigma_k}\ar@<-0.0ex>[d] & N^k \ar@<-0.0ex>[d] \ar@{-->}[r]^{\sigma_k^{-1}} & N_k \ar[d] \\ 
          \St(n, R[t, t^{-1}]) \ar@<-0.0ex>[r]_{\beta_{\delta_k}} \ar[d]_{\pi} & \St(n, R[t, t^{-1}]) \ar@<-0.0ex>[r]_{\beta_{\delta_k^{-1}}} \ar[d]_{\pi} & \St(n, R[t, t^{-1}]) \ar[d]_{\pi} \\
          \GL(n, R[t, t^{-1}]) \ar@<-0.0ex>[r]_{(-)^{\delta_k}} & \GL(n, R[t, t^{-1}]) \ar@<-0.0ex>[r]_{{}^{\delta_k}(-)} & \GL(n, R[t, t^{-1}])}
\end{equation}
For $n\geq 5$ these morphisms have already been constructed by Tulenbaev, see \S~3 of~\cite{Tu83}.
Thus, in the sequel we are mainly interested in the case $n=4$.
To simplify notation, we limit ourselves to constructing the map $\sigma_1$, so everywhere in the sequel we assume that $k=1$.

Of course, the map of right conjugation with $\delta_1$ induces the obvious one-to-one set-theoretic correspondence between $\mathcal{X}_1$ and $\mathcal{X}^1$. 
However, there seems to be no reasonable description of the complete set of relations between the elements $\mathcal{X}_1$ that would allow one to describe $N_1$
 as a group given by generators and relations.
Thus, in order to define $\sigma_1$ we will use the alternate description of $N^1$ given in terms of ``another presentation''.
We start with the following simple observation.
\begin{lemma} \label{lem:n1-decomp} There is an isomorphism $N_1 \cong N_{1,0} \rtimes P_1^-(R)$, where $N_{1,0}$ denotes the subgroup $\St(n, R[t], tR[t])$
 and $P_1^-(R)$ is the subgroup of $\St(n, R)$ generated by $x_{ij}(\xi)$ with $i\neq 1$.
\end{lemma}

Recall that the universal property of semidirect products asserts that for any group $H$ acting on a group $N$ via $\phi \colon H \to \Aut(N)$ and any group homomorphisms
 $f_N\colon N \to G$, $f_H\colon H \to G$ satisfying 
\begin{equation} \label{eq:coherence-condition} f_N(\phi(h)(n)) = f_H(h) f_N(n) f_H(h)^{-1},\ n\in N,\ h\in H,\end{equation} there exists a unique map $f\colon N \rtimes H \to G$
 extending $f_N$ and $f_H$. 
Thus, by the above lemma, in order to construct the map $\sigma_1$ we need to construct the following two maps and verify that they satisfy~\eqref{eq:coherence-condition}:
\[ (\sigma_1)_{P_1^-(R)} \colon P_1^-(R) \to N^1, \ \ (\sigma_1)_{N_{1,0}} \colon N_{1,0} \to N^1.\]

It is easy to define the first map $(\sigma_1)_{P_1^-(R)}$.
Using the decomposition $P_1^-(R) \cong U^-_1 \rtimes L_1$ where 
\[U^-_1 = \langle x_{i1}(\xi) \mid i\neq 1,\ \xi\in R \rangle \text{ and } L_1 = \langle x_{ij}(\xi) \mid i,  j \neq 1,\ \xi\in R\rangle \]
we apply the universal property of semidirect products once again 
 and define $(\sigma_1)_{P_1^-(R)}$ by requiring that it acts identically on $L_1$ 
and acts on elements of $U^-_1$ via the formula $(\sigma_1)_{P_1^-(R)}(x_{i1}(\xi))= x_{i1}(t\xi)$.

Before we proceed with the construction of the second map $(\sigma_1)_{N_{1,0}}$ we need to recall several technical
 constructions pertaining to ``another presentation'' of the group $N_{1,0}$.

For $u \in R^n$ we denote by $D(u)$ the subset of $R^n$ consisting of all $v$ satisfying $u^tv = 0$ and having at least two zero entries.
Recall from 3.2 of~\cite{Ka77} that for every $u, v, w \in R[t]^n$ such that $u^t \cdot v = 0$ there
 is a decomposition of $(w^t u) \cdot v$ into a sum of elements of $D(u)$, called {\it canonical decomposition}:
\setcounter{equation}{1}
\renewcommand{\theequation}{\arabic{equation}}
\begin{equation} \label{eq:canonical} (w^tu) \cdot v=\sum_{i<j}u_{ij} c_{ij}(v, w),\end{equation}
where $u_{ij}=e_iu_j-e_ju_i,$ and $c_{ij}(v, w)=v_iw_j-v_jw_i.$

Let $A$ be a commutative ring and $I$ be a splitting ideal of $A$.
Let $u, v \in A^n$ be such that $u^t v = 0$ and assume, moreover, that either $u$ or $v$ has at least one zero entry.
Recall from 3.8, 3.10 of~\cite{Ka77} that under these assumptions one can define certain element
 $x(u, v)$ of $\St(n, A)$ such that $\pi(x(u, v)) = t(u, v) = 1 + uv^t$.
The following properties of $x(u, v)$ are standard (cf.~\cite[Lemma~1.1]{Tu83}).
\begin{lemma} \label{lem:xsmall-properties}
\begin{lemlist}
 \item \label{item:xsmall-scalar} If $v$ or $w$ has at least two zero components, then 
 \begin{equation}\nonumber x(v, wa) = x(va, w).\end{equation}
 \item \label{item:xsmall-additivity} If $w_1$ and $w_2$ have at least one common zero coordinate then
 \begin{equation}\nonumber x(v, w_1)x(v, w_2) = x(v, w_1+w_2).\end{equation}
 \item \label{item:xsmall-commute} If $v$, $v'$ are simultaneously orthogonal to $w$ and $w'$ and the elements $w, w'$
  both have at least two zero components then $x(v, w)$ and $x(v', w')$ commute.
\end{lemlist}
\end{lemma}

Let $u \in E(n, A)e_1$, $v \in I^n$ be such that $u^tv = 0$.
Recall from \S~4 of~\cite{LS17} that one can define the following elements of $\St(n, A, I)$:
\begin{equation} \label{eq:sigma-definition}
 F(u, v) = \prod\limits_{i=1}^r x(u,  v_i),\ \
 S(v, u) = \prod\limits_{i=1}^r x( v_i, u).
\end{equation}
Here $\{v_r\}$, $r=1,\ldots,N$ is any collection of elements of $I^n \cap D(u)$ such that $\sum_{r=1}^N v_r = v$.
Since $u$ is unimodular, such a collection always exists by~\eqref{eq:canonical}, moreover
 $F(u, v)$ and $S(u, v)$ do not depend on its choice. %TODO: Give reference

Recall from~\cite{LS17} that it is possible to give a presentation of the relative Steinberg group $\St(n, A, I)$ formulated in terms of elements $F(u, v)$, $S(v, u)$.
More formally, \cite[Proposition 3.10]{LS17} asserts that $\St(n,\,A,\,I)$ admits presentation with the set of generators
\begin{multline*} \{F(u,\,v)\mid u\in E(n,\,A)e_1,\ v\in I^n,\ u^tv=0\}\, \cup \\ \cup \{S(u,\,v)\mid u\in I^n,\ v\in E(n,\,A)e_1,\ u^tv=0\}\end{multline*}
subject to relations
\setcounter{equation}{0}
\renewcommand{\theequation}{R\arabic{equation}}
\begin{align}
&F(u,\,v)F(u,\,w)=F(u,\,v+w), \label{add4}\\
&S(u,\,v)S(w,\,v)=S(u+w,\,v), \label{add5}\\
&F(u,\,v)F(u',\,v')F(u,\,v)\inv=F(t(u,\,v)u',\,t(v,\,u)\inv v'), \label{conj3} \\
&F(ge_1,\,g^*e_2a)=S(ge_1a,\,g^*e_2),\ \text{for all}\ a\in I,\, g\!\in \E(n, A). \label{coinc} \end{align}
Here $g^*$ denotes the contragradient matrix, i.\,e. $g^* = {g^{-1}}^t$ and $e_i$ is the standard basis vector of $A^n$.

\setcounter{equation}{3}
\renewcommand{\theequation}{\arabic{equation}}

To define the map $(\sigma_1)_{N_{1,0}}$ we will invoke the above presentation in the special case $A=R[t]$, $I=tR[t]$.
Indeed, define $\sigma_1$ on $N_{1,0}=\St(n, R[t], tR[t])$ by the following identities: 
\begin{equation}
 \sigma_1(F(u, v)) = \prod\limits_{i=1}^r x(\delta_1^{-1} \cdot t u, \delta_1 \cdot t^{-1} v_i),\ \
 \sigma_1(S(v, u)) = \prod\limits_{i=1}^r x(\delta_1^{-1} \cdot v_i, \delta_1 \cdot u).
\end{equation}

\begin{lemma} \label{lem:cor-conj}
 \begin{lemlist}
   \item \label{item:correctness} The elements $\sigma_1(F(u, v))$ and $\sigma_1(S(v, u))$ are well-defined, i.\,e. they do not depend on the choice of decomposition for $v$.
   \item \label{item:conjugation} For any $g \in N^1$ one has 
      \begin{equation}
          \nonumber {}^g\left(\sigma_1(F(u, v))\right) = \sigma_1\left(F(g' \cdot u, {g'}^{*} \cdot v)\right),\ \ 
                    {}^g\left(\sigma_1(S(u, v))\right) = \sigma_1\left(S(g' \cdot u, {g'}^{*} \cdot v)\right),                    
      \end{equation}
      where $g' = {}^{\delta_1}\pi(g).$
 \end{lemlist}
\end{lemma}
\begin{proof}[Short proof]
 The first assertion can be easily verified using the canonical decomposition and~\cref{lem:xsmall-properties}
  (compare with Tulenbaev's argument after~\cite[Lemma~1.1]{Tu83}).

 The second assertion can be proved similarly to~\cite[3.14]{Ka77}, (cf. also with the proof of~\cite[Lemma~4.4d]{LS17}).
\end{proof}
\begin{proof}[Long proof]
 Let us verify the first assertion. Let $v = \sum_r v^r$ be a decomposition as above. 
 Since each $v^r$ is orthogonal to $u$ we can write the canonical decomposition  
 $v^r = \sum_{i<j} u_{ij} c_{ij}(v^r, w)$, moreover, $\sum_{r} c_{ij}(v^r, w) = c_{ij}(v, w)$.
 \cref{lem:xsmall-properties} allows us to rewrite the right-hand side of~\ref{eq:sigma-definition} in such a way that 
  it does not depend on the choice of decomposition for $v$:
 \begin{multline}
  \prod\limits_r x(\delta_1^{-1} ut, \delta_1 v^rt^{-1}) = \prod\limits_{r}\prod\limits_{i<j} x(\delta_1^{-1} ut, \delta_1 u_{ij}c_{ij}(v^r, w)t^{-1}) = \\ 
   = \prod\limits_{i<j} x(\delta_1^{-1} ut, \delta_1 u_{ij}c_{ij}(v, w)t^{-1}). 
 \end{multline}

 We only verify the second assertion for $F(u, v)$  and only in the special case $g = x_{hk}(\xi) \in \mathcal{X}^1$, $g' = t_{hk}(\xi') \in \pi(\mathcal{X}_1)$.
 Choose $w\in R^n$ so that $w^t u = 1$ and write $v = \sum_{1\leq i<j\leq n} u_{ij} c_{ij}$, where $u_{ij}$ and $c_{ij} = c_{ij}(v, w)$ are as in~\eqref{eq:canonical}. 
 We can write ${}^g\left(\sigma_1\left(F(u, v)\right)\right)$ as follows:
 \begin{equation} \nonumber
   \prod\limits_{1\leq i<j\leq n} x(\pi(g) \cdot \delta_1^{-1} \cdot t u, \pi(g)^* \cdot \delta_1 \cdot t^{-1} u_{ij}c_{ij})
                                   = \prod\limits_{1\leq i<j\leq n} x(\delta_1^{-1} \cdot g'\cdot t u, \delta_1 \cdot {g'}^* \cdot t^{-1} u_{ij}c_{ij}).
 \end{equation} 
 Now for every factor in the right-hand side we do the following.
 If $\{i, j\} = \{h, k\}$ or $\{i, j\} \cap \{h, k\} = \emptyset$ we leave the term unchanged.
 Otherwise, if, $|\{i, j\} \cap \{h, k\}| = 1$,  say $j = h$, $i\neq k$, we further decompose it as follows:
  \begin{equation} \nonumber
    x(\delta_1^{-1} \cdot t u', \delta_1 \cdot {g'}^* \cdot u_{ij} (c_{ij} t^{-1})) = 
     x(\delta_1^{-1} \cdot u' t, \delta_1 \cdot u'_{ij} (c_{ij} t^{-1})) \cdot 
     x(\delta_1^{-1} \cdot u' t, \delta_1 \cdot u'_{ki} (c_{ij} t^{-1})).
  \end{equation}
  Here $u' = g'  u = t_{jk}(\xi') \cdot u$.
  Notice that ${g'}^*u_{ij} = t_{kj}(-\xi') \cdot u_{ij} = e_iu_j - e_ju_i + e_k\xi'u_i =u'_{ij}+u'_{ki}\xi'$.

  Thus, we arrive at a product $\prod_{s} x(\delta_1^{-1} \cdot u't, \delta_1 \cdot v'_s t^{-1}) $, where $v'_s \in D(u')$ and
   $\sum v'_s = {g'}^*v$. Clearly, this product equals $\sigma_1\left(F(g' \cdot u, {g'}^{*} \cdot v)\right)$ by definition.
\end{proof}

\begin{cor}
 The map $\sigma_1$ preserves relations~\eqref{add4}--\eqref{conj3}.
\end{cor}
\begin{proof}
 For~\eqref{add4}--\eqref{add5} this is an immediate corollary of the first assertion of the previous lemma.
 For~\eqref{conj3} this follows from the second part of the lemma applied in the special case $g = \sigma_1(F(u, v)) \in N^1$.
\end{proof}

It remains to verify that $\sigma_1$ preserves~\eqref{coinc}.
Let us do this at first in the special case when $g^*$ belongs to the subset $G_0 = H_{12}(R) \cdot U^+_1(R) \cdot U^-_1(R)$.
Here we denote by $H_{12}(R)$ the subgroup of $T(n, R)$ generated by semisimple root elements $ h_{12}(\xi)$, $\xi \in R^*$.

Since the only nonzero components of $u = g^* \cdot e_2$ are $u_1$ and $u_2$ we can present $v = g \cdot e_1$ as a sum $v' + v''$, where 
 $v'_i=0$, $i>2$ and $v''_1=v''_2=0$.
Since $v', v'' \in D(u)$ and $u \in D(v)$, from the definition of $\sigma_1$ we obtain that:
\begin{multline} \label{eq:special-case}
 \sigma_1(S(g \cdot e_1a, g^* \cdot e_2)) = x(\delta_1^{-1} \cdot v'a, \delta_1\cdot  u) \cdot x(\delta_1^{-1}\cdot v''a, \delta_1 \cdot u) = \\
  = x(\delta_1^{-1} va, \delta_1 \cdot u) = x(\delta_1^{-1} vat^{-1}, \delta_1 \cdot ut) = \sigma_1(F(g \cdot e_1, g^* \cdot e_2a)) 
\end{multline}

To obtain the assertion in the general case notice that for $R$ local the group $\E(n, R)$ admits the following decomposition:
\[\E(n, R) = \mathrm{EP}_1(R) \cdot H_{12}(R) \cdot U^-_1(R) \cdot U^+_1(R).\]
This can be either obtained as a corollary of Gauss decomposition or can be proved by a direct calculation. %TODO: Give reference

Applying transpose-inverse automorphism $(-)^*$ to the above decomposition and invoking~\cref{lem:n1-decomp} we obtain that $\E(n, R[t]) = \pi(N_1) \cdot G_0$.
Thus, every $g \in \E(n, R[t])$ can be factored as $\pi(n) \cdot h$ for some $n\in N_1$ and $h \in G_0$.
Since $\pi(n) = {}^{\delta_1} \pi(n')$ for some $n' \in N^1$ it remains to apply~\cref{item:conjugation} and~\eqref{eq:special-case} to obtain the required assertion:
\begin{multline} \nonumber \sigma_1(F(g \cdot e_1, g^* \cdot e_2a)) = {}^{n'}(\sigma_1(F(h \cdot e_1, h^* \cdot e_2a))) = \\
                           = {}^{n'}(\sigma_1(S(h \cdot e_1a, h^* \cdot e_2))) = \sigma_1(S(g \cdot e_1a, g^* \cdot e_2)). \end{multline}

Thus, we have completed the construction of the map $(\sigma_1)_{N_{1,0}}$.
It is not hard to verify that the condition~\eqref{eq:coherence-condition} is satisfied with these definitions,
 therefore, we have also completed the construction of the map $\sigma_1$.
